---
title: "SysBio Day 1 Hands On"
output:
  pdf_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup, include=F}
knitr::opts_chunk$set(echo=T)
par(las=1)
library(knitr)
library(kableExtra)
```


## Note

Here you'll find the code for the exercises of day 1. Do not worry if your code
from following the live coding session is not exactly the same, there are always
several ways to perform the same operation. If you find trouble trying to run
anything when coding, remember that the documentation and/or Google are your
best friends.

## 1) Installing packages
Dependencies for today's exercises include:

| Package | source | Description |
| :----: | :---: | :-------------------: |
| `biomaRt` | Bioconductor | Get molecular annotation for features (gene, proteins, etc) |
| `Biobase` | Bioconductor | Handle Expression-Sets (gene expression data) |
| `GEOquery` | Bioconductor | Get data from GEO |
| `qusage` | Bioconductor | Read GMT files |

Depending on your R version, bioconductor packages are installed in different
ways.
You can check you R version using `R.version$version.string`

```{r ins_dep1, eval=F, results=F, message=F}
# For R version < 3.5
source("https://www.bioconductor.org/biocLite.R")

biocLite("biomaRt")
biocLite("Biobase")
biocLite("GEOquery")
biocLite("qusage")
```

```{r ins_dep2, eval=F, results=F, message=F}
# For R version > 3.5
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("biomaRt")
BiocManager::install("Biobase")
BiocManager::install("GEOquery")
BiocManager::install("qusage")
```

Let's now load the libraries:

```{r load_lib, results=F, message=F}
library("biomaRt")
library("Biobase")
library("GEOquery")
library("qusage")
```

## 2) Loading data set from GEO
You can visit the GEO entry for `GSE116436` at 

> https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE116436

`GSE116436` is the GEO series of the whole data set. It comprehends
transcriptome profiling upon **60 cell lines using 15 anti-cancer agents** with
different time points (2, 6, 24 hrs) and drug concentrations (incl. null
concentration, i.e. control). A short description and experimental overview are
shown. For further details, you can read the original article:

> Monks A et al. The NCI Transcriptional Pharmacodynamics Workbench: A Tool to
Examine Dynamic Expression Profiling of Therapeutic Response in the NCI-60 Cell
Line Panel. Cancer Res 2018 Dec 15;78(24):6807-6817. PMID:
[30355619](https://www.ncbi.nlm.nih.gov/pubmed/30355619)

You can also click on one sample (transcriptome profile) to see
**sample metadata**.

> For instance,
[GSM3231645](https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSM3231645)

Importantly, this **data set is large (7.5k samples)**.
Thus the GEO series is stratify in subseries by each drug:

| GEO accession | Drug |
| :--- | :----------------------- |
| GSE116437 |	5-Azacytidine |
| GSE116438 |	bortezomib |
| GSE116439 |	cisplatin |
| GSE116440 |	dasatinib |
| GSE116441 |	doxorubicin |
| GSE116442 |	erlotinib |
| GSE116443 |	geldanamycin |
| GSE116444 |	gemcitabine |
| GSE116445 |	lapatinib |
| GSE116446 |	paclitaxel |
| GSE116447 |	sirolimus |
| GSE116448 |	sorafenib |
| GSE116449 |	sunitinib |
| GSE116450 |	topotecan |
| GSE116451	|vorinostat |

Let's start with one subset of the data to understand how it works.

We choose the `GEO accession number` of one subserie. For instance, we choose
lapatinib (GSE116445).

We use the following command line to download the data
```{r pressure, cache=TRUE, results=F, message=F}
lapatinib <- getGEO("GSE116445", destdir="/tmp", AnnotGPL=TRUE)[[1]]
```

## 3) Exploring the data object

After this, we have a R object named `lapatinib`. This object contains the gene
expression data and sample metadata. To explore this object, we can use:

```{r expl_dim_eset, results=F, message=F}
class(lapatinib)

# It is a Expression-Set, which is a special object to store 'expression data'
is(lapatinib)

# What is the dimension of this bunch of data?
dim(lapatinib)

# Which attributes does the object have?
attributes(lapatinib)
```

### How do the features and samples look like?
```{r expl_dim_eset2, results=F, message=F}
# Features = probes
head(featureNames(lapatinib), 5)

# Samples = transcriptome profiles (under certain experimental conditions)
head(sampleNames(lapatinib), 5)

```

### Wrap-up
You can see that both `Features` and `Samples` are described throughby
**identifiers**...

In order to analyse and interpret this data, we need to map this identifiers to
the genes and experimental conditions of each sample.

## 4) Subsetting the data

For the day 2 practice, we have subset the data set in a specific manner.
We have choosen the relevant profiles as those perturbed with the highest drug
concentration and the untreated samples (as control) and latest time point,
expecting to see the strongest patterns. To do so, we first get the sample
metadata for these two features: drug concentration and time point

```{r sample_metadata, results=F, message=F}
# the sample metadata contains the drug concentrations within tags
sample_tags = as.character(lapatinib$title)

## Drug concentration
# Extract the part describing drug concentration
drug_conc = sapply(sample_tags, function(tag){
  strsplit(tag, split="_")[[1]][3]
  })
# Replace nanoMolar with empty string
drug_conc = gsub("nM$", "", drug_conc)
# Convert to numbers
drug_conc = as.numeric(drug_conc)
# Max conc
drug_conc.max = max(drug_conc)
# Min conc
drug_conc.min = min(drug_conc)

## Time point: basically the same
drug_tpoint = sapply(sample_tags,function(tag){
  strsplit(tag, split="_")[[1]][4]
  })
drug_tpoint = gsub("h$", "", drug_tpoint)
drug_tpoint = as.numeric(drug_tpoint)
drug_tpoint.max = max(drug_tpoint)
```

Finally we subset the data

```{r subset_prof, results=F, message=F}
selected_samples = (drug_conc == drug_conc.max | drug_conc == drug_conc.min) & drug_tpoint == drug_tpoint.max
lapatinib = lapatinib[, selected_samples]

dim(lapatinib)
```

## 5) Access to the gene expression and annotation data

Expression-Sets are special objects to store gene expression data and metadata
associated to both genes and samples. All this information is store in a unique
`R object`.

**The Gene Expression data**: Gene expression data is store as a quantitative
matrix. This matrix contains measurements of the gene-level expression.
**Metainformation for samples and features**: Two `data.frames` contains
metadata for samples and features.

To access to the gene expression data, we use `exprs()`.

```{r exprs_access, results=F, message=F}
e_data = exprs(lapatinib)
head(e_data)
```

`Expresion-Set` objects store the metadata information within the object
itself.
To access to this information, there are two main functions from `Biobase`
package:

* `pData()` : Access to phenotypeData. The phenotypeData is the sample metadata
described in the GEO entry of each sample.
* `fData()` : Access to featureData. The featureData is complementary
information for each one of the features (probes).

Let's try them out!

### Sample metadata
```{r pdata_access, results=F, message=F}
p_data = pData(lapatinib)
head(p_data)
```

### Feature metadata
```{r fdata_access, results=F, message=F}
f_data = fData(lapatinib)
head(f_data)
```

## 6) Identifier conversion

Microarrays are High-Throughput platforms throughby gene-level expression is
measured using one or a few probes for the same gene.

For instance, we could know which platform was used in our dataset:
```{r plat, results=F, message=F}
# GEO platform identifier
annotation(lapatinib)
# Or using the sample metadata
unique(pData(lapatinib)$hyb_protocol)
```

In order to create a fit matrix of Genes x Samples, we are going to collapse
the intensity of one or more probes into one single value that represents the
expression of the gene.
For this, we need to create a vector that map each probe to its corresponding
gene

```{r match, cache=T, eval=F, results=F, message=F}
# Loading Mart object for Ensembl Homo sapiens data
mart = useMart('ENSEMBL_MART_ENSEMBL', 'hsapiens_gene_ensembl',
               host = "www.ensembl.org", ensemblRedirect=F)

query = featureNames(lapatinib)
matching_table = getBM(c('affy_hg_u133a_2', 'external_gene_name'),
                       filters='affy_hg_u133a_2', values=query, mart=mart)
colnames(matching_table) = c('probe', 'gene')
```

### Workaround (if Ensembl servers are down)

```{r eval=F, results=F, message=F}
# Quick and dirty
matching_table = data.frame(1:length(f_data$ID))
matching_table$probe = f_data$ID
# Getting unique IDs (first gsymbol)
matching_table$gene = gsub('[///].*',  '', f_data$`Gene symbol`)
matching_table$gene = gsub('[-].*',  '', matching_table$gene)
matching_table = matching_table[, -1]
```

### Renaming columns
```{r eval=F, results=F, message=F}
# Convert columns names to something readable
our_columns <- colnames(e_data)
our_columns <- p_data[our_columns, 'title']
colnames(e_data) <- our_columns
```

### Collapsing probe intensities to gene-level expression
```{r col_expr, eval=F, results=F, message=F}
# Transform into a data.frame
df = as.data.frame(e_data)

# Matching Identifiers
gsymbols = as.character(sapply(rownames(df), function(x){
  aux = matching_table[matching_table$probe == x, 'gene'][1]
  }))

# We add another column to the data.frame which describe the corresponding gene
df$gene = gsymbols

# Aggregate values for the same gene by the average
df2 = aggregate(. ~gene, data=df, mean)

# Replace row.names with genes
rownames(df2) = as.character(df2$gene)

# Removing unmapped GeneSymbols
df2 = df2[!rownames(df2) == '', ]

# Remove the column with the genes
df2 = df2[, which(colnames(df2) != "gene")]
```

## 7) Saving the data
```{r save_data, eval=F, results=F, message=F}
# Saving the expression data
write.csv(df2, 'my_data.csv')

# Let us save the annotations for the upcoming days
write.csv(p_data, 'my_p_data.csv')
```

## 8) Retrieving pathway annotation

Let's take as an example a curated gene set (C2) from [MSigDB](http://software.broadinstitute.org/gsea/msigdb/).

Download the `.gmt` file from the Biocarta curated pathways and subset the
list of genes corresponding to the MAPK pathway (`BIOCARTA_MAPK_PATHWAY`).
> Alternatively, you can search directly on the website and dowload only that
> specific list of genes([here](http://software.broadinstitute.org/gsea/msigdb/cards/BIOCARTA_MAPK_PATHWAY.html))

```{r load_gset, eval=F, results=F, message=F}
gsets = read.gmt('c2.cp.biocarta.v7.1.symbols.gmt')
gset = gsets[['BIOCARTA_MAPK_PATHWAY']]
```

## 9) Obtaining PPI network

Now we will download the protein-protein interaction network from [OmniPath](http://omnipathdb.org/interactions?genesymbols=1).

```{r opath, eval=F, results=F, message=F}
url = 'http://omnipathdb.org/interactions?genesymbols=1'
opath = unique(read.table(url, sep='\t', header=T)[, c(3, 4)])

# Subsetting the PPI to members of the pathawy
pathway = opath[opath[, 1] %in% gset & opath[, 2] %in% gset, ]
```

## 10) Mapping expression to pathway

Finally, we extract the median expression of all nodes in our pathway and save
them in a file as well as the network itself.

```{r map_exp_path, eval=F, results=F, message=F}
unique_genes = unique(c(as.character(pathway[, 1]),
                        as.character(pathway[, 2])))

measurements = data.frame()

# Extracting  median of measurements of pathway members
for(g in unique_genes){
  measurements[g, 1] = rowMedians(as.matrix(df2[g, ]))
}

# Storing the results
write.table(pathway, 'pathway.txt', sep='\t', quote=F, row.names=F)
write.csv(measurements, 'measurements.csv', quote=F)
```


