---
title: "Introduction to footprint analysis"
author: "Aurelien Dugourd"
date: "6/19/2019"
output:
  html_document: default
  pdf_document: default
---

#Introduction

In this practical we will run a transcription factor analysis and a pathway analysis using footprint based methods.

The goal here is to play around with the parameters of the analysis to get a good sense of how to interpret the results.

#Getting everything we need

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=4, fig.width=12}
library(readr)

lung_vs_renal_ttop <- as.data.frame(read_csv("Dropbox/course_HD/data/lung_vs_renal_ttop.csv")) #Got this from the limma analysis of day3

load("~/Dropbox/course_HD/dorothea_viper.Rdata") #This is generated with the df_to_viper_regulon() function of omictoolstest (see github) with DOROTHEA's v2 regulons

progeny_coef_matrix <- as.data.frame(read_csv("Dropbox/course_HD/model_NatComm+14_human.csv")) #Got this from Christian

source("~/Dropbox/course_HD/support_functions.R")
```

#Checking the differential analysis result
```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
library(ggplot2)
library(ggrepel)

volcano_nice(lung_vs_renal_ttop, FCIndex = 2, IDIndex = 1, pValIndex = 5, nlabels = 30, label = T)
```

#Classic pathway analysis with GSEA
```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
library(qusage)
library(fgsea)

KEGG <- read.gmt("~/Dropbox/course_HD/gene_set_collection/c2.cp.kegg.v6.2.symbols.gmt") #Get this directly from msigdb

gene_diff_expression <- lung_vs_renal_ttop$t
names(gene_diff_expression) <- lung_vs_renal_ttop$Gene

pathway_analysis <- as.data.frame(fgsea(pathways = KEGG, stats = gene_diff_expression, nperm = 1000, minSize = 5, gseaParam = 1))
pathway_analysis <- pathway_analysis[order(pathway_analysis$padj, decreasing = F),]
```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
top_pathways <- as.data.frame(pathway_analysis[c(1:10),5])
row.names(top_pathways) <- as.character(pathway_analysis[1:10, 1])

library(pheatmap)

pheatmap(top_pathways, cluster_cols = F)
```
Now, try to change the gseaParam value. How does this affect your data, and why.

#Transcription factor analysis
```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
library(viper)

#Prepare the measurments for viper
viper_expression <- lung_vs_renal_ttop$t
names(viper_expression) <- lung_vs_renal_ttop$Gene

###This is just to handle a bug in viper
viper_expression <- as.data.frame(viper_expression)
viper_expression$dummy <- viper_expression$viper_expression

###COmpute TF activity scores
TFactivity <- as.data.frame(viper(eset = viper_expression, regulon = dorothea_viper, minsize = 2, adaptive.size = F, eset.filter = T))
names(TFactivity) <- "activity"
```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=5, fig.width=10}
TFactivity_top <- TFactivity
TFactivity_top$TF <- row.names(TFactivity_top)
TFactivity_top <- TFactivity_top[abs(TFactivity_top$activity) > 3,]

ggplot(TFactivity_top, aes(x = TF, y = activity)) + geom_bar(stat="identity") + theme_minimal()
```

Now, try to change the minsize paramater and eset.filter and see how it affects the results. How does the eset.filter parameter affect the interpretation of the result ?

#Progeny analysis
```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
progeny_pathway_scores <- runProgenyFast(lung_vs_renal_ttop[,c(1,4)], weight_matrix = progeny_coef_matrix, k = 1000)
names(progeny_pathway_scores) <- "score"
progeny_pathway_scores$pathway <- row.names(progeny_pathway_scores)
```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=5, fig.width=10}
ggplot(progeny_pathway_scores, aes(x = pathway, y = score)) + geom_bar(stat="identity") + theme_minimal()
```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
progeny_pathway_scores_pval <- progeny_pathway_scores
progeny_pathway_scores_pval$pval_lower <- pnorm(progeny_pathway_scores$score, lower.tail = T)
progeny_pathway_scores_pval$pval_upper <- pnorm(progeny_pathway_scores$score, lower.tail = F)
```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}
progeny_res_df <- list()
for(i in 1:100)
{
  progeny_res_df[[i]] <- runProgenyFast(lung_vs_renal_ttop[,c(1,4)], weight_matrix = progeny_coef_matrix, k = 1000)
}
progeny_res_df <- do.call(cbind,progeny_res_df)
names(progeny_res_df) <- paste("Run_",c(1:100),sep="")

print(mean(apply(progeny_res_df,1,sd)))
```
Now play around with the number of permutations and see how it affect the stability of the result. Why does it affect it this way ?
```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}

```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}

```

```{r, message=FALSE, warning=FALSE,results='hide',fig.keep='all', fig.height=10, fig.width=10}

```
